---
title: "Project 1 Predictive Modeling"
output: html_notebook
---

# Introduction

During this project, we will analyse two data sets in order to obtain more insights about them and predict some objective variable. First, we will approach the problem with a Linear Regression model, then, with the second data set, we will try to model a Generalised Linear Model. These are the data sets used:

-   Airbnb: <https://www.kaggle.com/datasets/dipeshkhemani/airbnb-cleaned-europe-dataset>

-   Data scientist jobs: <https://www.kaggle.com/datasets/nikhilbhathi/data-scientist-salary-us-glassdoor>

# Linear Regression

## Data preparation

For the first part of the assignment we will be using the Airbnb data set. It consist of information about the Airbnb stays in some European cities. The variables are the following:

-   City: Amsterdam, Athens, Barcelona, Berlin, Budapest, Lisbon, Paris, Rome or Vienna
-   Price: continuous numeric variable
-   Day: WeekDay or Weekend
-   Room Type: Entire home/apt, Private room or Shared room
-   Shared Room: binary variable
-   Private Room: binary variable
-   Person capacity: integer from 2 to 6.
-   Superhost: binary variable
-   Multiple Rooms: binary variable
-   Business: binary variable
-   Cleanliness Rating: integer from 2 to 10.
-   Guest satisfaction: continuous numeric variable
-   Bedrooms: integer from 0 to 10
-   City center(km): continuous numeric variable
-   Metro distance(km): continuous numeric variable
-   Attraction index: continuous variable measuring the attraction to the Airbnb by the costumers
-   Normalised Atraction index: same but normalised
-   Restraunt Index: continuous variable measuring the restaurant performance index
-   Normalised Restraunt Index: same but normalised

The goal would be to predict the Price variable in terms of the others. First, it is always a good idea to separate from the beginning the training set (what the tool is going to see) from the testing set (used only to validate predictions)

```{r}
library(tidyverse)
library(MASS)
library(caret) #ML tools
library(e1071) 
par(mar = c(2, 2, 2, 2))
# Loading and preparing data
Airbnb <- read.csv("Airbnb_data.csv")
Airbnb
```

```{r}
# split between training and testing sets
spl = createDataPartition(Airbnb$Price, p = 0.8, list = FALSE)  # 80% for training

AirbnbTrain = Airbnb[spl,]
AirbnbTest = Airbnb[-spl,]

str(AirbnbTrain)

```

Once separated, let's clean and understand the training set. First, we check any missing values. There is any missing value.

```{r}
any(is.na(AirbnbTrain))
```

Now we evaluate the numerical variables, having some statistics such as minnimum, maximum, quantiles mean and median.

```{r}
summary(AirbnbTrain)

```

From the data obtained we can obtain the following insigths:

##### Price

The maximum of price seems to be too large for the mean and median. We can get a better look with a boxplot:

```{r}
boxplot(AirbnbTrain$Price)
```

Although the maximum is an outlier, it is not alone, as we can observe many other outliers that could be representing expensive Airbnb. In the case the maximum would be alone, we could get rid of it, but in this case this can be meaningful data.

##### Attraction.Index and Restraunt.Index

As we have the normalize columns, we have redundant information, so we will only keep those.

```{r}
AirbnbTrain <- subset(AirbnbTrain, select = -c(Attraction.Index, Restraunt.Index ))

```

The rest of numerical variables seem fine with their ranges, so no changes are needed. Let's now evaluate the categorical variables:

```{r}
table(AirbnbTrain$City)
table(AirbnbTrain$Day)
table(AirbnbTrain$Room.Type)
table(AirbnbTrain$Shared.Room)
table(AirbnbTrain$Private.Room)
table(AirbnbTrain$Superhost)
table(AirbnbTrain$Multiple.Rooms)
table(AirbnbTrain$Business)

```

The variable Room type has the same information than Private Room and Shared room, so we will erase the last two.

```{r}
AirbnbTrain <- subset(AirbnbTrain, select = -c(Private.Room, Shared.Room ))

```

Finally, the variable Superhost is as False/True, we will code as binary the False/True options.

```{r}
AirbnbTrain$Superhost<-as.numeric(as.logical(AirbnbTrain$Superhost))

```

## Explanatory analysis

In order to better comprehend our data, we need to plot it. Concretely, we will have in mind plots respect to the objective variable, the price, for trying to visualize a model for predicting it.

```{r}
ggplot(AirbnbTrain, aes(Price)) + geom_density(fill="lightblue") + xlab("Price") + ggtitle("Price distribution")

## from the statistics obtained before we know price has no 0 values as minimum is 34, so we can apply the logarithm to better visualize it


ggplot(AirbnbTrain, aes(log(Price))) + geom_density(fill="lightblue") + xlab("log(Price)") + ggtitle("Price distribution")

mean(log(AirbnbTrain$Price))
exp(mean(log(AirbnbTrain$Price)))


```
As can be observed, Price is positive asymmetric, as most of the data is located at the left of the distribution. In order to make the data more symmetric, we need to take the logarithm.
The log(Price) seems to follow a Gaussian distribution with mean 5.36 (214.2107 €). This is beneficial in terms of future prediction of the target, as it is symetrical there are no much extreme values and the regresion would make better predictions. However, still after the transformation the data are not fully centered, but are slightly  shifted to the left. 

After observing the response variable alone, we can approximate some regressions with plots. We now have in mind the Prices in terms of the city, in order to better understand the data.

```{r}
ggplot(AirbnbTrain, aes(log(Price))) + geom_density(aes(group=City, colour=City, fill=City), alpha=0.1) + ggtitle("Price distribution")
```

Most of the cities have the same distribution, however we can observe how Berlin has the highest prices while Amsterdam the lowest.

Some of the variables that we may think are important to determine the price of an Airbnb are: distance to metro, distance to city center, number of bedrooms, restaurant and atraction indexes. Let's observe them.

```{r}
ggplot(AirbnbTrain, aes(x=City.Center..km., y=log(Price), color=City)) + geom_point(alpha=0.8) + ggtitle("Price vs City Center distance") 

```

```{r}
ggplot(AirbnbTrain, aes(x=Metro.Distance..km., y=log(Price), color=City)) + geom_point(alpha=0.8) + ggtitle("Price vs Metro distance")
```

As observed, the most expensive Airbnb are the ones really close to the metro and center (there are no points at the rigth upper corner).

```{r}
ggplot(AirbnbTrain, aes(x = Bedrooms, y = log(Price), fill = City)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  ggtitle("Price vs Number of bedrooms")
```

```{r}
table(AirbnbTrain$Bedrooms)
```

We could say that surprisingly, Airbnb with higher number of bedrooms are cheaper than the ones with lower number. However, we don't have enought information about the number of bedrooms higher than 4, so it is difficult to make any assumptions. From 0 to 3, the ones with 2 bedrooms seems the most expensive in general.

```{r}
ggplot(AirbnbTrain, aes(x=Normalised.Attraction.Index, y=log(Price), color=City)) + geom_point(alpha=0.8) + ggtitle("Price vs Attraction Index") 
```

```{r}
ggplot(AirbnbTrain, aes(x=Normalised.Restraunt.Index, y=log(Price), color=City)) + geom_point(alpha=0.8) + ggtitle("Price vs Restraunt Index") 
```

Finally, the indexes did not provide as much information. But, focusing on the Restaurant index we could visualize a kind of linear tendency with lower prices on the ones with lower index and higher prices close to higher index rating. However, is still difficult to make assumptions without an appropiate model. In the following sections this relationship would be better explained.

### Correlations

The most important term when evaluating a data set for predictions are the correlations. Which are the variables that are more correlated with our objective variable? First, we would focus on isolated correlations:

```{r}
## Moving the objective function to the last position
AirbnbTrain<-AirbnbTrain[, c(1:(2 - 1), (2 + 1):ncol(AirbnbTrain), 2)]

```

```{r}
corr_delay <- sort(cor(AirbnbTrain[,c(4:15)])["Price",], decreasing = T)
corr=data.frame(corr_delay)
ggplot(corr,aes(x = row.names(corr), y = corr_delay)) + 
  geom_bar(stat = "identity", fill = "lightblue") + 
  scale_x_discrete(limits= row.names(corr)) +
  labs(x = "", y = "Price", title = "Correlations") + 
  theme(plot.title = element_text(hjust = 0, size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Assuming you have loaded the required libraries
library(ggcorrplot)

# Compute the correlation matrix
cor_matrix <- cor(AirbnbTrain[, c(4:15)])

# Create a correlation matrix plot using ggcorrplot
ggcorrplot(cor_matrix,
           colors = c("lightblue", "white", "lightcoral"),  # Define color scale
           lab = FALSE,  # Show correlation values
           lab_size = 1,  # Set label size
           title = "Correlation Matrix")

# If you still want to plot the correlations with the response variable "Price"
corr_delay <- sort(cor_matrix["Price", ], decreasing = TRUE)
corr_data <- data.frame(variable = names(corr_delay), correlation = corr_delay)

# Plot the correlations
ggplot(corr_data, aes(x = variable, y = correlation)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(x = "", y = "Correlation with Price", title = "Variable Correlations") +
  theme(plot.title = element_text(hjust = 0.5, size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))


```

A priori, the only variables helpful for performing a linear regression predictor are Normalised Atraction Index, Normalised Restraunt Index, number of Bedrooms and Person Capacity. However, is important to recall that we are plotting the individual correlations, maybe if we add information between variables we could get higher correlations, and therefore have better predictions.

### Simple linear regression: interpreting the betas

Before developing a complex model for prediction, we will try to create a simpler one where we avoid colinearities and better understand the betas. Therefore, we first create a simple linear regressor with the higher correlation variable: Normalised Atraction Index.

```{r}
simple_linFit <- lm(log(Price) ~ Normalised.Attraction.Index, data=AirbnbTrain)
summary(simple_linFit)
```

The simple linear regression model is defined as: $log(Price) = \beta_0 + \beta_1 N.Attraction.Index= 5.01 + 0.03 N.Attraction.Index$

where $\beta_1$ explains that a 1 unit increase in the normalized attraction index would increase in 0.03 % the log of the Price. We can trust this model because the information gain is larger than the noise, as the p_value close to 0 confirms.

Furthermore, focusing on the global information, the residual standard error, although is larger than the local, is still not that high. The residual standard error is higher than the local because there is not much information gain from the model, however, we can trust the model as the information that $\beta_1$ gives us has small standard error.

Finally, the $R^2$ tell us about the information gain we get from the model. Almost 20%, while the remaining 80% is noise. In general, this is a difficult application, as the most correlated variable is only giving us 20% of the information, so it would be difficult to obtain high $R^2$ values on the test set. However, we always need to compare with a benchmark for better conclusions (on the following sections).

To better comprehend how this model approaches the price prediction we will plot the regression line:

```{r}
# Scatter plot

ggplot(AirbnbTrain, aes(x = Normalised.Attraction.Index, y = log(Price))) +
  geom_point(alpha = 0.8, color= 'lightblue') +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  ggtitle("Price vs Attraction Index regression")
```

After understanding the model, let's evaluate how it predicts (this result will be later compared with the benchmark and the complex linear regression model).

```{r}
pred_simple= exp(predict(simple_linFit, newdata= AirbnbTest))

#Let's know validate the predictions, calculating R^2 in the testing set

R2_simple= cor(pred_simple, log(AirbnbTest$Price))^2
R2_simple
```


As expected, $R^2$ on the testing set is smaller than on the training set (remember it was 0.2). Always is more difficult to predict future values than present ones.

Finally, we can perform some diagnosis of our simple model to obtain more conclusions:

- Residuals vs Fitted Values: this plot show as how our model is not homoscedastic, that is, he variance of the residuals is not constant with the Price, but it increases (negatively) for higher prices. This can be explained by the non symmetric distribution and the outliers.

- Q-Q Residuals: the residuals are not normally distributed, as the outliers in the tails of the plot suggest issues with extreme residuals.

- Scale-Location Plot: again this plots confirm us our model is not homocedastic, as there is a clear deviation on the line.


In general, we can say our model is not good for our objective.
```{r}
par(mar = c(2, 2, 2, 2))

plot(simple_linFit, pch=23 ,bg='orange',cex=2)
```

## Predictive analysis

After understanding better how our data is constructed and being defined a simple regression model, we will know try to develop the best model for prediction. But first, we need to define the benchmark, from where we have to improve.

### Benchmark

In our case, as we are dealing with numerical continuous target, we would consider the benchmark as predicting the mean (again using the logarithmic transformation). This would be similar to the following model: $log(price)= \beta_0 \times 1$

```{r}
benchFit <- lm(log(Price) ~ 1, data=AirbnbTrain)
pred_bench <- exp(predict(benchFit, newdata=AirbnbTest))
```

As on the mean predictor the $R^2$ can not be computed, we will use the Mean Squared Error for comparison:

```{r}
#MSE comparison: my model vs benchmark
MSE_simple=sqrt(mean((pred_simple - log(AirbnbTest$Price))^2))
MSE_bench= sqrt(mean((pred_bench - log(AirbnbTest$Price))^2))
MSE_simple
MSE_bench
```

Clearly, our simple model is worse than the naive model, so we need to improve it.

### Complex model (without interactions)

As a first approach to find our perfect model, we would fit a model with all the variables we have. To do it, we first must change the format of the variable 'Superhost' on the test set for non finding programmatical errors when calculating the predictions.

```{r}
AirbnbTest$Superhost<-as.numeric(as.logical(AirbnbTest$Superhost))

```

```{r}
all_linFit <- lm(log(Price) ~ ., data=AirbnbTrain)

# Predict on the test set
pred_all <- exp(predict(all_linFit, newdata = AirbnbTest))
  
  # Calculate MSE
MSE_all <- sqrt(mean((pred_all - log(AirbnbTest$Price))^2))
R2_all= cor(pred_all, log(AirbnbTest$Price))^2
summary(all_linFit)
R2_all
MSE_all
```

First, if we compare the $R^2$ values in the training and in the testing sets with the simple model, before we had  0.2 and 0.016, respectively. The value for the training set has increased until 0.66 by using all the variables, and the value on the testing set has increased on 10% until 0.16. Of course, this complex model is better than the simpler one.

However, when comparing with the MSE, we can see how this model has the worst score with 268.068, followed by the simple with 240, and again the benchmark is the best one with 208.068.

In order to try to improve this metrics, now we need to take into account interactions between variables. 

### Complex model (with interactions)

We first start by doing some feature selection, as does not have any sense to add all the features because we will be adding more noise. Therefore, we would try to find the best combination of the most correlated variables: Normalised.Attraction.Index, Normalised.Restraunt.Index, Bedrooms and Person.Capacity.

```{r}
Y <- log(AirbnbTrain$Price)

#variables_to_add= list("Normalised.Attraction.Index", "Normalised.Restraunt.Index", "Bedrooms", "Person.Capacity")

#variables_to_add= list( "Bedrooms", "Person.Capacity", "Normalised.Attraction.Index", "Normalised.Restraunt.Index")

#variables_to_add= list("Bedrooms", "Normalised.Attraction.Index", "Normalised.Restraunt.Index",  "Person.Capacity")

variables_to_add= list("Person.Capacity", "Normalised.Attraction.Index", "Normalised.Restraunt.Index", "Bedrooms" )

# Create an empty data frame to store models and MSE
mse_df <- data.frame(MSE = numeric(length(variables_to_add)), R2= numeric(length(variables_to_add)), Variables = character(length(variables_to_add)), stringsAsFactors = FALSE)

models <- list()

# Fit models and calculate MSE
for (i in seq_along(variables_to_add)) {
  formula <- as.formula(paste("Y ~ ", paste(variables_to_add[1:i], collapse = " + ")))
  model <- lm(formula, data = AirbnbTrain)
  
  # Store the model
  models[[i]] <- model
  
  # Predict on the test set
  pred <- exp(predict(model, newdata = AirbnbTest))
  
  # Calculate MSE
  mse <- sqrt(mean((pred - log(AirbnbTest$Price))^2))
  
  R2= cor(pred, log(AirbnbTest$Price))^2

  # Store MSE and variables in the data frame
  mse_df[i, "MSE"] <- mse
  mse_df[i, "R2"] <- R2
  mse_df[i, "Variables"] <- paste(variables_to_add[1:i], collapse = ", ")
}

# Find the model with the lowest MSE
best_model_index <- which.max(mse_df$R2)
best_model <- models[[best_model_index]]

# Print results
cat("Best model includes variables:", mse_df$Variables[best_model_index], "\n")
cat("R2 of the best model:", mse_df$R2[best_model_index], "\n")

```

```{r}
best_linFit <- lm(log(Price) ~ City*(Normalised.Attraction.Index + Normalised.Restraunt.Index)^2 , data=AirbnbTrain)

# Predict on the test set
pred <- exp(predict(best_linFit, newdata = AirbnbTest))
  
  # Calculate MSE
MSE_best <- sqrt(mean((pred - log(AirbnbTest$Price))^2))
R2_best= cor(pred, log(AirbnbTest$Price))^2
R2_best
MSE_best
```

```{r}
day_linFit <- lm(log(Price) ~  City, data=AirbnbTrain)

# Predict on the test set
pred <- exp(predict(day_linFit, newdata = AirbnbTest))
  
  # Calculate MSE
MSE_day <- sqrt(mean((pred - log(AirbnbTest$Price))^2))
R2_best= cor(pred, log(AirbnbTest$Price))^2
R2_best
MSE_day


```



### Prediction Intervals

# Generalised Linear Models

## Data preparation

## Model development and interpretation

## Predictive performance
